---
title: "My Expenses Data Exploration"
output:
  html_document: default
  html_notebook: default
---
This tiny dataset which contains about more than 1000 rows with 9 variables each has information about my expenses from April 2015 to till date. This dataset is obtained from an andriod app "Walnut", which is a expense tracker.   

## 1. Loading The Datadet And Required Imports
```{r message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(plotly)
setwd("E:/MOOC")
expensesTemp <- read.csv("walnut.csv",stringsAsFactors=FALSE)
```

## 2. Dimensions Of The Dataset
```{r}
dim(expensesTemp)
```

## 3. Description of the variables and their types.
```{r}
str(expensesTemp)
```

## 4. Data Cleaning 
As part of data cleaning, lets remove some unnecessary columns and transform some variables like date and amount from chr variables to date and number variables. The Account variable contains some sensitive information like last 4 digits of my debit and credit cards, so I am removing that variable from the dataframe.  

Apart from removing some columns, I have added columns like year and month to track year-wise and month-wise expenses.
```{r}
expensesTemp <- subset(expensesTemp,select = -c(NOTE,TIME,ACCOUNT,TAGS))
expensesTemp$DATE <- as.Date(expensesTemp$DATE,format='%d-%m-%y')
expensesTemp$MONTH <- as.factor(format(expensesTemp$DATE,'%b'))
expensesTemp$YEAR<- as.factor(format(expensesTemp$DATE,'%Y'))
expensesTemp$PLACE <- as.factor(expensesTemp$PLACE)
expensesTemp$EXPENSE <- as.factor(expensesTemp$EXPENSE)
expensesTemp$CATEGORY <- ifelse(expensesTemp$CATEGORY == "","UNKNOWN",expensesTemp$CATEGORY)
expensesTemp$CATEGORY <- as.factor(expensesTemp$CATEGORY)
expensesTemp$AMOUNT <- as.numeric(gsub(",","",expensesTemp$AMOUNT))
str(expensesTemp)
```

## 5. Expense Variable And Total Expenses:
The expense variable has mainly two values, YES and NO, if the values is YES, the record can be counted as expense.
```{r}
summary(expensesTemp$EXPENSE)
```
Now Lets transform our expenses dataset so that it includes only those records, that can be counted as expenses.
```{r}
expenses <- subset(expensesTemp,expensesTemp$EXPENSE == "Yes")
```
  
#### The Total Amount Of Money Spend Till Date:
```{r}
sum(expenses$AMOUNT)
```
#### The Total Amount Spent By Year:
```{r}
with(expenses,by(AMOUNT,YEAR,sum))
```
```{r}
expenses_by_year <- expenses %>% 
  group_by(YEAR) %>%
  summarize(total=sum(AMOUNT),n=n())

ggplot(data=expenses_by_year,aes(x=YEAR,y=total))+
  geom_bar(stat="identity",color="black", width=0.3,fill="#F4511E")+
  geom_text(aes(label=total), vjust=-1.2) +
  scale_y_continuous(breaks = seq(0,400000,50000),limits =c(0,400000))+
  labs(y="AMOUNT SPENT",x="DIFFERENT YEARS",title="YEAR V/S SPENDS")+
  theme_bw()+
  theme(axis.text=element_text(size=10),axis.title=element_text(size=13))+
  theme(plot.title = element_text(hjust = 0.5,size = 14,face = "bold"))+
  ggsave(filename = "outputs/fig-1.png",width = 5,height = 4)  
```

From here on, I will be considering the expenses of the 2017 only, as they are better categorized then the expenses of 2015 and 2016. 

## 6. Category Varaiable :
The category variable captures the different type of categories I spend in.
```{r}
levels(expenses$CATEGORY)
```
Now lets calculate category wise spends:
```{r warning=FALSE}
expenses <- subset(expenses,YEAR == "2017")
expenses_by_category <- expenses %>% 
  group_by(CATEGORY)%>%
  summarize(total=sum(AMOUNT),n = n())%>%
  arrange(desc(total))

expenses_by_category
```

```{r fig.width = 10}
ggplot(data=arrange(expenses_by_category,total),aes(x=reorder(CATEGORY,total),y=total))+
  geom_bar(stat="identity",color="black",fill="#F4511E",width = 0.6)+
  geom_text(aes(label=total), hjust=-0.2) +
  coord_flip()+
  scale_y_continuous(breaks = seq(0,101000,25000),limits = c(0,109000))+
  labs(y="AMOUNT SPENT",x="DIFFERENT CATEGORIES",title="CATEGORIES V/S SPENDS")+
  theme_bw()+
  theme(axis.text=element_text(size=10),axis.title=element_text(size=13))+
  theme(plot.title = element_text(hjust = 0.5,size = 14,face = "bold"))+
  ggsave(filename = "outputs/fig-2.png",width = 8,height = 5)  
```

## 7. Month Varaible:
In this section I will be exploring my month wise expenses.
```{r}
expenses$MONTH <- factor(expenses$MONTH,ordered = TRUE, levels = c("Jan","Feb","Mar","Apr","May","Jun","Jul","Aug","Sep","Oct","Nov","Dec"))
expenses_by_month <- expenses %>% 
  group_by(MONTH) %>%
  summarize(total = sum(AMOUNT),n=n())

expenses_by_month
```

```{r fig.width = 10}
ggplot(data=expenses_by_month,aes(x=MONTH,y=total))+
  geom_bar(stat="identity",color="black",fill="#F4511E",width=0.6)+
  geom_text(aes(label=total), vjust=-1.4) +
  labs(y="AMOUNT SPENT",x="DIFFERENT MONTHS",title="SPENDS V/S MONTH")+
  scale_y_continuous(breaks = seq(0,65000,10000),limits = c(0,62000))+
  geom_hline(yintercept = 20000,linetype=2,size=0.9,color="black")+
  theme_bw()+
  theme(axis.text=element_text(size=12),axis.title=element_text(size=13))+
  theme(plot.title = element_text(hjust = 0.5,size = 14,face = "bold"))+
  ggsave(filename = "outputs/fig-3.png",width = 10,height = 5)  
```

Now lets combine month variable with category variable to explore the month wise category spends.

```{r}
expenses_by_month_category <- expenses %>% 
  group_by(MONTH,CATEGORY) %>%
  summarize(total = sum(AMOUNT),
            n= n())
expenses_by_month_category
```

```{r message=FALSE, warning=FALSE,fig.width = 10, fig.height = 10}
p <- ggplot(data=expenses_by_month_category,aes(x=MONTH,y=total,fill=CATEGORY))+
  geom_bar(stat = "identity",color="black",position = "stack")+
  labs(y="AMOUNT SPENT",x="DIFFERENT MONTHS",title="CATEGORY SPENDS MONTH WISE")+
  scale_y_continuous(breaks = seq(0,60000,5000))+
  scale_fill_manual(values = c("#98FB98","#4B0082","grey","pink","green",
                               "yellow","red","yellow","#6A5ACD","#00BFFF","#FF00FF","darkblue","darkred"))+
  theme_bw()+
  theme(axis.text=element_text(size=12),axis.title=element_text(size=14,face="bold",hjust = -1))+
  theme(plot.title = element_text(hjust = 0.5,size = 14,face = "bold"),axis.title=element_blank(),axis.ticks.y=element_blank())+
  ggsave(filename = "outputs/fig-4.png",width = 7,height = 5)  

p
```

To see a interactive plot of month wise category spends uncomment the code below and run the code chunk.
```{r message=FALSE, warning=FALSE,fig.width = 10, fig.height = 15}

#ggplotly(p)

```

## 8. Misc Facts

#### A). Total Number Of ATM Visits And Amount Drawn:
```{r}
totalAtmVisits_2017 <- subset(expensesTemp,expensesTemp$PLACE == "ATM" & expensesTemp$YEAR == "2017")
sprintf("Total Number Of Atm Visits: %s", nrow(totalAtmVisits_2017))
sprintf("\nTotal Amount of Money Withdrawn: %s",sum(totalAtmVisits_2017$AMOUNT))
```

Now lets compare the above 2017 ATM stats to 2016 ATM stats to see whether **Demonitisation** played any role in number of ATM withdrawals. 
```{r}
totalAtmVisits_2016 <- subset(expensesTemp,expensesTemp$PLACE == "ATM" & expensesTemp$YEAR == "2016")
sprintf("Total Number Of Atm Visits: %s", nrow(totalAtmVisits_2016))
sprintf("Total Amount of Money Withdrawn: %s",sum(totalAtmVisits_2016$AMOUNT))
```

The above stats clearly show that demonetization played a role in decreasing my number of ATM withdraws and increasing my card spends.

#### B). Total Number Of Cab Rides Taken Till Date.
```{r}
cabRides <- sum(with(expensesTemp,PLACE == "OLA" |  PLACE == "OLA CABS" | PLACE == "UBER" ))
sprintf("Total Number Of Cab Rides Taken: %s",cabRides)
```

#### C). Total Number Of Tea Cups Consumed.
```{r}
teaCups <- sum(with(expensesTemp,PLACE == "HATTI KAPPI"| PLACE == "HATTI KAAPI" | PLACE == "TEA" ))
sprintf("Total Number Tea Cups Consumed: %s",teaCups)
```

#### D) Total Number Of Visits To UDUPI FOODS.
```{r}
udupiFoods <- sum(with(expensesTemp,PLACE == "UDIPI FOODS" | PLACE == "UDUPI FOODS"))
sprintf("Total Number Of Visits To UDUPI FOODS: %s",udupiFoods)
```