---
title: "My Expenses Data Exploration"
output:
  html_document: default
  html_notebook: default
---
## 1. Loading The Datadet
```{r message=FALSE, warning=FALSE}
setwd("E:/MOOC")
expenses <- read.csv("walnut.csv",stringsAsFactors=FALSE)
```

## 2. Dimensions Of The Dataset
```{r}
dim(expenses)
```

## 3. Description of the variables and thier types.
```{r}
str(expenses)
```

## 4. Data Cleaning 
As part of data cleaning, lets remove some unnecessary columns and transform some variables like date and amount from chr variables to date and number variables.  
```{r}
#expenses <- subset(expenses,select = -c(NOTE,TIME))
expenses$DATE <- as.Date(expenses$DATE,format='%d-%m-%y')
expenses$MONTH <- as.factor(format(expenses$DATE,'%b'))
expenses$PLACE <- as.factor(expenses$PLACE)
expenses$ACCOUNT <- as.factor(expenses$ACCOUNT)
expenses$EXPENSE <- as.factor(expenses$EXPENSE)
expenses$CATEGORY <- ifelse(expenses$CATEGORY == "","UNKNOWN",expenses$CATEGORY)
expenses$CATEGORY <- as.factor(expenses$CATEGORY)
expenses$TAGS <- as.factor(expenses$TAGS)
expenses$AMOUNT <- as.numeric(gsub(",","",expenses$AMOUNT))
str(expenses)
```

## 5. Expense Variable And Total Expenses:
The expense varaiable has mainly two values, YES and NO, if the values is YES, the record can be counted as expense.
```{r}
summary(expenses$EXPENSE)
```
Now Lets transform our expenses dataset so that it includes only those records, that can be counted as expenses.
```{r}
expenses <- subset(expenses,expenses$EXPENSE == "Yes")
```
  
#### The Total Amount Of Money Spend Till Date:
```{r}
sum(expenses$AMOUNT)
```

## 6. Category Varaiable :
The category variable captures the diiferent type of categories I spend in.
```{r}
levels(expenses$CATEGORY)
```
Now lets calculate category wise spends:
```{r}
with(expenses,by(AMOUNT,CATEGORY,sum))
```















